---
title: "Basic programming I"
author: "Susy Echeverría-Londoño and Drew Kerkhoff"
output:
  html_document:
    highlight: tango
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_float: yes
---
``This document is a work in progress, please give us feedback``

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functions

Functions allows you to perform a specific task in a more powerful and general way. When you need to run the same thing repeatedly, the most efficient and reproducible way is to write a function. Functions will avoid incidental mistakes when you copy and paste code in different parts of your script. 

Let's look at the following example of code which rescales each column of a data frame to a range from 0 to 1

```{r}
 df<-data.frame(
   a=rnorm(10),
   b=rnorm(10),
   c=rnorm(10)
 )

df$Rescale_a<-(df$a-min(df$a,na.rm=TRUE))/(max(df$a,na.rm=TRUE)-min(df$a,na.rm=TRUE))
df$Rescale_b<-(df$b-min(df$b,na.rm=TRUE))/(max(df$b,na.rm=TRUE)-min(df$b,na.rm=TRUE))
df$Rescale_c<-(df$c-min(df$c,na.rm=TRUE))/(max(df$c,na.rm=TRUE)-min(df$c,na.rm=TRUE))

df$Rescale_a
```

# Writing a function

Functions need to be written to be reusable. For that we need: 
- Clear variable names
- Consistent indentation and line spacing
- Good commenting

## Basic structure

```{r}
function_name <- function(inputs) {
  output_value <- do_something(inputs)
  return(output_value)
}
```

Let's try our first function called rescale01

```{r}
rescale01<-function(x){
  
  rescale_value<-(x-min(x, na.rm=TRUE))/(max(x,na.rm=TRUE)-min(x,na.rm=TRUE))
  return(rescale_value)
  
}

rescale01(x=df$a)

rescale01(df$a)

df$Rescale_a<-rescale01(df$a)

```

We can combine functions to do larger tasks in two ways. In this case, we can use the function `range()` which gives us the minimum and maximum values directly

```{r}

rescale01<-function(x){
  
  ran<-range(x, na.rm=TRUE)
  rescale_value<-(x-ran[1])/(ran[2]-ran[1])
  
  return(rescale_value)
  
}

rescale01(df$a)

df$Rescale_a<-rescale01(df$a)
```

## Walk through function execution

- Call function `rescale01`
- Assign the vector `df$a` to the `x` argument (or input)
- Calculate the range of the vector
- Using the min and max values rescale the vector and store the new values in the `rescale_value` object
- Send the `rescale_value` object back as output
- Store it in `df$Rescale_a`

You can also write the function in a simpler way. 

```{r}
rescale01<-function(x){
  
  (x-min(x, na.rm=TRUE))/(max(x,na.rm=TRUE)-min(x,na.rm=TRUE))

}
```

## Challenge

1. Write a function that calculates the volume of a plant using lenght, width, height as input arguments. `volume=length * width * height`
2. Calculate the volume for the following plants


| Plants | length (in) | width (in) | height (in) |
|--------|-------------|------------|-------------|
| A      | 20.32       | 40.64      | 50.80       |
| B      | 25.40       | 38.10      | 60.96       |
| C      | 25.4        | 50.8       | 127         |

3. Calculate a function that converts inches in centimeters (_there are 2.54 cm in one in_). Use that function to transform the lenght, width, height plant values. 

4. Create a new function that calculates the volume of a plant in `cm^3` using a combination between the functions in 1 and 3.  


# Conditionals

Conditionals evaluate whether a logical statement is either `TRUE` or `FALSE`. Conditional statements use operators such as `==`, `!=`, `<`,`>`, `<=`, `>=`. You can combine these with `&` (and) or `|` (or). You can also use some predefine functions such as `identical()`

```{r}
20>7

"A"=="B"

"A"!="B"

identical("C","D")
```

## Challenge

_Taken from data carpentry[http://www.datacarpentry.org/semester-biology/exercises/Making-choices-choice-operators-R/]_
 
1. Create the following variable

```{r}
w <- 10.2
x <- 1.3
y <- 2.8
z <- 17.5
dna1 <- "attattaggaccaca"
dna2 <- "attattaggaacaca"
```

2. Use them to print whether or not the following statements are `TRUE` or `FALSE`

- w is greater than 10
- w + x is less than 15
- x is greater than y
- 2 * x + 0.2 is equal to y
- dna1 is the same as dna2
- dna1 is not the same as dna2
- w is greater than x, and y is greater than z
- x times w is between 13.2 and 13.5
- dna1 is longer than 5 bases (use nchar() to figure out how long a string is), or z is less than w * x

## Floating point

```{r}
x<-sqrt(2)^2
x

x==2
```

Floating points are unexpected results from computer arithmetics in numbers that contain very small trailing digits. 

```{r}
# Shows the number in a C-style format
formatC( x, format='f', digits=20)
```

To avoid floating-point issues, use the function `dplyr::near()` or round the values of the variable

```{r message=FALSE, warning=FALSE}
require(dplyr)

near(x, 2)

round(x)==2
```

## if statements

`if` statements allows you to conditionally execute code

```{r eval=FALSE}
if(condition){
  # code executed when condition is TRUE
}else{
  # code executed when condition is TRUE
}

```

### One condition example

```{r}
Size_tree<-function(x){
  
  if(x>5){
    print("Large tree")
  }else{
    print("Small tree")
  }

}

Size_tree(100)
Size_tree(3)
```

### Multiple conditions

```{r}
Size_tree<-function(x){
  
  if(x>5){
    print("Large tree")
  }else if(x<=5 & x>3 ){
    print("Medium tree")
  }else{
    print("Small tree")
  }

}

Size_tree(100)
Size_tree(5)
Size_tree(2)
```

```{r}

in_to_cm<-function(x){
  x/2.54
}

PlantVolume_in_m<-function(l, w, h, units="cm"){
  
  if(units=="in") {
    
    new_l<-in_to_cm(l)/100
    new_h<-in_to_cm(h)/100
    new_w<-in_to_cm(w)/100
    volume <- new_l*new_h*new_w
    
  }else if (units=="cm"){
    
    new_l<-l/100
    new_h<-h/100
    new_w<-w/100
    volume <- new_l*new_h*new_w
    
  }else if (units=="m"){
    
    volume <- l*h*w
    
  }else{
    
    message("Use either m, cm or in")
    volume<-NULL
  }
  
  return(volume)
  
}

PlantVolume_in_m(0.10,0.11,0.12,units="m")
PlantVolume_in_m(25.4,27.94,30.48,units="in")
PlantVolume_in_m(10,11,12,units="cm")
PlantVolume_in_m(10,11,12,units="km")
```
  
## Challenge

Go to this [link](http://www.datacarpentry.org/semester-biology/exercises/Making-choices-complete-the-code-R/) and complete the assignment from Data carpentry for Biologist website

# Sources
- Hadley Wickham and Garrett Grolemund. 2017. [R for Data Science](http://r4ds.had.co.nz/functions.html): Import, Tidy, Transform, Visualize, and Model Data (1st ed.). O'Reilly Media, Inc.

- Data carpentry for Biologist. [Functions in R](http://www.datacarpentry.org/semester-biology/materials/functions-R/)





