---
title: "Basic programming II"
author: "Susy Echeverría-Londoño and Drew Kerkhoff"
output:
  html_document:
    highlight: tango
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_float: yes
---
``This document is a work in progress, please give us feedback``

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(kableExtra)
```

# Merging data frames

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

## Data frames
```{r }
# 1. Traits data frame
sp_traits<-read.csv("./Data/species_traits.csv")

# 2. Biome distribution data frame
sp_biomes<-read.csv("./Data/species_biomes_distri.csv")

# 3. Growth form information
Growth_form<-read.csv("./Data/GrowthForm.csv")
```

## Exploring data

```{r}
# Exploring the dataframe
dim(sp_traits)
dim(sp_biomes)

# Alternative you can use View()
head(sp_traits)
head(sp_biomes)

# How many species are match between the two dataframes
table(sp_traits$species%in%unique(sp_biomes$species))

# We first need to make sure the species vector names have the same format
head(sp_traits$species)
head(sp_biomes$species)
```


## Matching variables and formats
```{r}
# Replacing spaces in species names
sp_traits$species<-gsub(" ","_",sp_traits$species)
head(sp_traits$species)

table(sp_traits$species%in%unique(sp_biomes$species))
table(unique(sp_biomes$species)%in%sp_traits$species)
```


## Merging data frames

```{r}
# ?merge
# When species columns have the same name
Trait_biomes_df<-merge(x=sp_traits,y=sp_biomes) 

head(Trait_biomes_df)

DF1<-merge(x=sp_traits,y=sp_biomes,
                       by.x="species",by.y="species")

DF2<-merge(x=sp_biomes,y=sp_traits,
                        by.x="species",by.y="species")

```

Is there a difference between data frame `DF1` and `DF2`?

## Challenge

Include the Growth form information into the `Trait_biomes_df` data frame. Remember to check first the format and names of the variables you want to match between the two data frames. 

```{r include=FALSE}
Growth_form$SPECIES_STD<-gsub(" ","_",Growth_form$SPECIES_STD)

tmp1<-merge(Trait_biomes_df,Growth_form,
            by.x="species",by.y="SPECIES_STD")

tmp2<-merge(Growth_form,Trait_biomes_df,
            by.x="SPECIES_STD",by.y="species")


Trait_biomes_df<-merge(Trait_biomes_df,Growth_form,
            by.x="species",by.y="SPECIES_STD")
```


# Functions practice

Let's recap the main structure of a function in R! 



```{r}
add<-function(x, y = 1){
  x + y
}

format(add)
body(add)
environment(add)

add(x=2)

add(x=4,y=6)
```

## Let's practice

Extracting the mean values of seed mass per family in Moist forest
```{r}
Trait_biomes_df %>% 
  filter(Biome=="Moist_Forest") %>% 
  group_by(FAMILY_STD) %>% 
  summarise(Mean_Trait=mean(Seed_mass, na.rm=TRUE))
```

### Challenge
1. Create a function that takes any biome name as input and display the average value of seed mass per family

2. Modify the function to calculate the average value of any trait

3. Complete the `----` in the following code to create a function that:

a. Takes as input the name of a biome
b. Evaluate if this biome has species that belong to the family of true grasses (i.e., Poaceae)
c. Return the list of true grass species
d. If there are no grass species, returns a message "No grasses found in this biome"

```{r eval=FALSE, include=TRUE}

Biome_grasses<-function(biome){

  Biome_df<-
  Trait_biome_GF %>% 
    filter(Biome==____)
  
  if("Poaceae"%in%___){
   
    Biome_df %>% 
      filter(FAMILY_STD=="Poaceae") %>% 
      pull(species)
    
  }else{
    
    print(___)
    
  }  

}

```




