---
title: "Data Visualization - Exercise 4: Plotting Multiple Data Sets and Fitting Linear Models\\\n Solution"
author: "Luna L Sanchez Reyes"
date: "2023-03-06"
output: 
  pdf_document: default
  html_document: 
    toc: true
    toc_float: true
    keep_log: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tinytex.clean = TRUE) # removes aux files after rendering to html
library(ggplot2)
library(readr)
```

There are 3 mains steps to solve this exercise.

### 1) Read the data in with quality control

```{r}
trees <- read_tsv("../data-raw/TREE_SURVEYS.txt",
               col_types = list(HEIGHT = col_double(),
                                AXIS_2 = col_double()))
acacia <- read.csv(file = "../data-raw/ACACIA_DREPANOLOBIUM_SURVEY.txt",
         sep = "\t",
         na.strings = "dead")
```

### 2) Quality Assurance

Visualize the data to assure it is good:

```{r}
str(trees)
```


```{r}
str(acacia)
```

### 3) Plot the two data sets on the same plot and fit linear models.

The trick to do this is to call the `ggplot()` function with no data set.
This allows us to provide a different data set for any and each function we call after.
In this way we can plot data from different data set unto the same plot.
The downside is that we have to specify the data set for **EVERY** layer that we want to add.
This is different from what happens when we specify the data within the base layer `ggplot()`.



```{r}
ggplot() +
  geom_point(data = trees, mapping = aes(x = CIRC, y = HEIGHT), color = "gray", alpha = 0.5) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(data = trees, mapping = aes(x = CIRC, y = HEIGHT), method = "lm", color = "black") +
  geom_point(data = acacia, mapping = aes(x = CIRC, y = HEIGHT), color = "red", alpha = 0.8) +
  geom_smooth(data = acacia, mapping = aes(x = CIRC, y = HEIGHT), method = "lm", color = "red") +
  labs(x = "Tree circumference", y = "Tree height")
```

