---
title: "Data Wrangling - Exercise 1: Data manipulation \\\n Solution"
author: "Luna L Sanchez Reyes"
date: "2023-03-02"
output: 
  html_document: 
    toc: true
    toc_float: true
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tinytex.clean = TRUE) # removes aux and log files after endering to html
# install.packages("dplyr")
library(dplyr)
```


### 1. Load the data set from the file `surveys.csv` into R using the function `read.csv()`.

```{r}
surveys <- read.csv(file = "../data-raw/surveys.csv")
str(surveys)
```

---

### 2. Use `select()` to create a new data frame object called `surveys1` with just the `year`, `month`,
   `day`, and `species_id` columns in that order.

- `select()` is from package `dplyr`, so we need to load the package, preferably in the "setup" R chunk, but can also be done here and commented out for knitting.

```{r}
# library(dplyr)
surveys1 <- select(surveys, year, month, day, species_id)
str(surveys1)
```

---

### 3. Create a new data frame called `surveys2` with the `year`, `species_id`, and weight **in kilograms** of each individual, with no null weights. 

Use `mutate()`, `select()`, and `filter()` with `!is.na()`. The weight in the table is given in grams so you will need to create a new column called "weight_kg" for weight in kilograms by dividing the weight column by 1000.

- The goal of this question is to make them realize that they have to create intermediate data frames.

First, filter Na values in weight:

```{r}
surveys_tmp <- filter(surveys, !is.na(weight))
str(surveys_tmp)
```

Second, create the new column with weight in Kg.
Overwrite the object.

```{r}
surveys_tmp <- mutate(surveys_tmp, weight_kg = weight/1000)
str(surveys_tmp)
```
Finally, select the columns that you want for the new `data frame`:

```{r}
surveys2 <- select(surveys_tmp, year, species_id, weight_kg)
str(surveys2)
```
---

### 4. Use the `filter()` function to get all of the rows in the data frame `surveys2` for the species ID "SH".

The goal of this point is to make them practice the function `filter()` and logical statements.

For next iterations of the course:

  - Tell them no to print the whole table in the knitted document. For this, there are a couple options:
    - You have to create a data frame object
    - You have to use r chunk options `results = 'hide`
  - Ask how many rows does the filtered table have?
  - Make them filter surveys1 instead, so you know that they did not overwrite it during question 2 and 3.

```{r}
surveys_filtered <- filter(surveys2, species_id == "SH")
str(surveys_filtered)
```
