---
title: "Plots and intro to spatial analysis"
author: "Susy Echeverría-Londoño and Drew Kerkhoff"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data visualizations in R

## 0. Data 

In this module, we will use the `` data object

```{r cars}

```

## Base graphics


```{r cars}
library(dismo)
```

# Intro to spatial analysis

## Download and clean geographical data
## Plot occurrences
## Create rasters 
## Calculate range size (squarekilometers)
## Calculate midpoint (lat and lon)
## Plot rasters (richness maps with letsR)


```{r pressure, echo=FALSE}
snake_sp<-gbif(genus="Acanthophis", species="rugosus", geo=TRUE)

## Cleaning data
snake_sp=subset(snake_sp, !is.na(lon) & !is.na(lat))

snakeDup=duplicated(snake_sp[, c("lon", "lat")])
snake_sp <-snake_sp[!snakeDup, ]


## Ploting 
library(maptools)
data(wrld_simpl)

plot(wrld_simpl, xlim=c(min(snake_sp$lon)-1,max(snake_sp$lon)+1), ylim=c(min(snake_sp$lat)-1,max(snake_sp$lat)+1), axes=TRUE, col="light yellow")
points(snake_sp$lon, snake_sp$lat, col="orange", pch=20, cex=0.75)

## Create presence absence matrix
xy<-data.frame(x=snake_sp$lon, y=snake_sp$lat)

# Presence/abscence raster of Solanum ----
snake_grids<-lets.presab.points(xy, snake_sp$species, resol=1)
snake_grids$Richness_Raster[snake_grids$Richness_Raster==0]<-NA

rangesize <- lets.rangesize(snake_grids,
                            coordinates = "geographic",units="squaremeter")

centroids <- lets.midpoint(snake_grids)

data_plot <- data.frame(centroids[, 2:3], "Range size" = rangesize)
g <- ggplot(data_plot, aes(x, Range_size))
g + geom_smooth() + geom_point() + labs(x = "Latitude(x)", y = "Range size")

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
